---
title: "FS Habitat"
author: "Bernard Romey"
date: "Tuesday, February 10, 2015"
output:
  pdf_document:
    fig_caption: yes
---
-------

My objective is to evaluate the possible use of canonical analysis of stream habitat data and fish abundance, or species diversity.  


```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=6, fig.height=6, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
```

```{r dta,echo=FALSE, results='hide', include=FALSE}
#Habitat
library(dplyr)
fs_hab <- read.csv("fs_habitat.csv")
hab <-fs_hab[c(1,6:9,12:14,16:18,20:21,26:28,30:31)]
hab <-na.omit(hab)
rm(fs_hab)
ave.hab <-hab%>%
  group_by(Site)%>%
  summarise_each(funs(mean))
  
hab.1 <-ave.hab[-c(1,2)]
hab.l <-scale(log(hab.1+1))
```

**Assumptions**

Due to the amount of missing data, I was unable to use all variables in the analysis (Table 1).  I ended up using the average for 26 study sites over multiple years for eighteen of the variables in the analysis.

The normality assumption for most variables after log transformation had a p-value > 0.05 (Failed to reject $H_o$ (Figure 1).  Only three of the variables were suspect with a p-value < 0.05 (Reject $H_o$).  

```{r box, fig.cap="Boxplot of all 18 variables (log transformed and standardized with Z-score) used in analysis"}
boxplot(scale(hab.1))
```


*Table 1.  Study site, Channel Type, reach length and area, and relative fish density(CT & DV).*


**Ordination**

PCA Biplot of eigenvectors for each variable showed that PCA reduced all variables to two principal components (Figure 2).  This was tested with a broken stick model to confirm the number of components selected by the model.  

The cosines of the angles between the descriptor-axes (eigenvector) is proportional to their covariances. For example, number of pools and pools per km are highly correlated (Figure 2).

Synthetic component one accounts for 37.9% of the variance in the data, while component two accounts for 20.4% of the variance in the data, for a total of 58.3% of the variance.  Loading (variance contribution) on component one are primarily made up of AvCBW, TotalArea, AvRPD, and Pool Spacing (-0.36, -0.33,-0.32, 0.29) that make up a pool distribution gradient.  Loading for component two are primarily D50,Pools.km, TKWD.km (0.37,-0.38, -0.29).  The gradient for component two is a little harder to interpret, as number of pools increase the D50 substrate decreases.  Since none of the loading are above 0.50 they are considered poor contributors.

```{r, echo=FALSE, fig.cap="Biplot of numeric habitat variables (Eigenvectors) indicating correlation among variables.  Scores(sites) showing euclidian distance to variable"}
require(MASS) #loads the PCA package
pca <- princomp(hab.l, cor=TRUE) #creates a PC matrix using the correlation matrix
biplot(pca, expand = 1.05,main = "Biplot", xlab = "Comp.1 (38.0%)", ylab = "Comp.2 (19.0%)")
```

Scree plot showing that approximately 50% of the variance in the data is represented in component one and two (Figure 3).

```{r Scree, fig.cap="Scree plot of variance contribution to component 1 & 2."}
plot(pca, main="Scree Plot") #Scree plot

```


The Shepard diagram shows the distortion of each observation when ordination reduces variables from multidimensional space to reduced space.  The farther from the y=x axis, the worse the distortion (Figure 4).  To what extent does the reduced space preserve the distance relationships among objects?

```{r shpPlot, fig.cap= "Shepard diagram showing reduced space distance distortion among observations"}
head(round(pca$scores,2))
euc<-dist(hab.l)#Computes the distance matrix in the log transformed data (multidimensional space)
euc.1<-dist(pca$scores[,c(1,2)]) #Computes the distance matrix in the PC matrix (reduced space)
plot(euc,euc.1,main="Shepards Diagram (PC=2)", xlab="Distance in Multidimensional space", ylab="Distance in Reduced space")   #Shepard diagram

#Farther from y=x the more distortion
```

**Discussion**

There are two main problems with this analysis: 1) The PCA components do a poor job of representing the variance in the data.  2) The distortion is high indicating that the euclidean distance is not well preserved among observations in the reduced space.  This indicates that the projection in reduced space does not account for a large fraction of the variance; the spaces are not similar.

Canonical analysis may be inadequate for evaluating fish abundance and diversity with this type of data, but it may be useful in evaluating other variables on a similar scale such as land use or watershed information.  


